### 2013-03-24 mcmc.tex
outline notes
nothing
what is a pdf, what is a sampling, badly normalized functions

Say you are in this situation.
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For our definition of a
  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
  series.}) $p(\data\given\pars)$.
% ************* BJB
% <Wild Subjective Bayesian>
% The "generative model" is a probabilistic description of your *prior
% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
% ************* Hogg
% That comment should go in the ``What is a Model?'' Chapter.
Furthermore, say you also can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement MCMC can give you
a fair sampling of the posterior pdf.
That is, you can run MCMC (for a very long time\note{Sometimes a very,
  very long time; we will return to that below.})  and you will be
left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setofall{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

@@

...What is a sampling, heurstically?.. Along the way, what is a pdf?..

If there is a pdf $p(x)$ for some quantity $x$, which can be a scalar
or a multi-element vector or list or blob, we can define
\emph{expectation values} $E[x]$ for $x$ or for any quantity that can
be expressed as a function $q(x)$ of $x$:
\begin{eqnarray}
E[x] &\equiv& \int x\,p(x)\,\dd x
\\
E[q] &\equiv& \int q(x)\,p(x)\,\dd x
\quad ,
\end{eqnarray}
where the integrals are implicitly definite integrals over the domain
of $x$ (all the parts of $x$ space in which $p(x)$ is finite) and the
integrals are multi-dimensional if $x$ is multi-dimensional.  These
expectation values are the mean values of $x$ and $q(x)$ under the
pdf.  A good sampling---really the definition of a good
sampling---makes the sampling approximation to these integrals
accurate.  With a good sampling $\setofall{x_k}_{k=1}^K$ the integrals
get replaced with sums over samples $x_k$:
\begin{eqnarray}
E[x] &\approx& \frac{1}{K}\,\sum_{k=1}^K x_k
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
\quad .
\end{eqnarray}
That is, a sampling is good when any expectation value of interest is
accurately computed via the sampling approximation.  Your conditions
on your sampling will depend on the expectations you want to compute,
and the accuracies you need.

@@

...Is there any way to reconstruct $p(x)$ from the samples
$\setofall{x_k}$?  In short, no, but in many cases of interest, sort-of!..

...You can't sample the likelihood!  You can only sample a posterior
pdf!.. why not?.. Give a sense of the difficulties you get into if you
fail here; like discussion with Sven Kreiss on 2013-02-14...

@@

% *********** BJB
% YES!!!! Although you can sample the likelihood if your samples are in
% data space. :)

@@

...Autocorrelation time is the premier tool.  It is not a panacea, and why?

@@

...How do you make it run fast?

...Acceptance ratio and ESJD and etc.

@@

...Gibbs sampling and interior marginalizations

>>>>

Say you are in this situation.
You have a huge blob of data $\data$ (think of this as a vector or
list or heterogeneous collection of observations\note{In general in
  this \documentname\ we will treat any variable as being capable of
  containing not just a value but some huge arbitrarily structured set
  of values in matrix, vector, or more complex form.}).
You also have a model sophisticated enough---a probabilistic,
generative model, if you will\note{For our definition of a
  \emph{model}, see [SOME OTHER NOTE] in this series.}---that, given a
setting of a huge blob of parameters\note{Again, the parameter
  blob $\pars$ should also be thought of as an immense, arbitrarily
  structured set or list of parameters, possibly even with no fixed
  dimension!} $\pars$, you can compute a pdf for data (or likelihood\note{We
  bash the terminology ``likelihood'' in [SOME OTHER NOTE] in this
  series.}) $p(\data\given\pars)$.
% ************* BJB
% <Wild Subjective Bayesian>
% The "generative model" is a probabilistic description of your *prior
% beliefs about what the data are going to be*. </Wild Subjective Bayesian>
% ************* Hogg
% That comment should go in the ``What is a Model?'' Chapter.
Furthermore, say you also can write down some kind of informative or
vague prior pdf $p(\pars)$ for the parameter blob.
If all these things are true, then---even if you can't compute
anything else---in principle a trivial-to-implement MCMC can give you
a fair sampling of the posterior pdf.
That is, you can run MCMC (for a very long time\note{Sometimes a very,
  very long time; we will return to that below.})  and you will be
left with a set of $K$ parameter-blob settings $\pars_k$ such that the
full set $\setof{\pars_k}_{k=1}^K$ constitutes a fair sampling from
the posterior pdf $p(\pars\given\data$).
% ************* BJB
% Notions of "sampling" and "Monte Carlo" are interesting. You might like to
% refer the interested reader to the paper "Monte Carlo is Fundamentally
% Unsound" by Tony O'Hagan.

@@

Heuristically, a sampling $\setof{x_k}_{k=1}^K$ from some pdf $p(x)$
is a set of $K$ values $x_k$ that are independent draws from the pdf.
Not being mathematicians, we don't know the precise definition of a
pdf\note{Seriously.  We don't.}, but for our purposes here a pdf is
any single-valued (scalar) function that is non-negative everywhere
(the entire domain of $x$), and obeys a normalization condition
\begin{eqnarray}
0 &\leq& p(x) \quad \mbox{for all $x$}
\\
1 &=& \int p(x)\,\dd x
\quad ,
\end{eqnarray}
where, implicitly, the integral is over the full domain of $x$.
Importantly, although $p(x)$ is single-valued, $x$ can be a scalar or
a multi-element vector or list or blob; it can be arbitrarily large
and complicated.
In data-analysis contexts, $x$ will often be the full blob of free
parameters in the model.

Given this pdf $p(x)$, we can define \emph{expectation values} $E[x]$
for $x$ or for any quantity that can be expressed as a function $q(x)$
of $x$:
\begin{eqnarray}
E[x] &\equiv& \int x\,p(x)\,\dd x
\\
E[q] &\equiv& \int q(x)\,p(x)\,\dd x
\quad ,
\end{eqnarray}
where the integrals are implicitly definite integrals over the domain
of $x$ (all the parts of $x$ space in which $p(x)$ is finite) and the
integrals are multi-dimensional if $x$ is multi-dimensional.  These
expectation values are the mean values of $x$ and $q(x)$ under the
pdf.  A good sampling---really the definition of a good
sampling---makes the sampling approximation to these integrals
accurate.  With a good sampling $\setof{x_k}_{k=1}^K$ the integrals
get replaced with sums over samples $x_k$:
\begin{eqnarray}
E[x] &\approx& \frac{1}{K}\,\sum_{k=1}^K x_k
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
\quad .
\end{eqnarray}
That is, a sampling is good when any expectation value of interest is
accurately computed via the sampling approximation.  The word
``accurately'' here translates into some kinds of theorems about
limiting behavior; the general idea is that the sampling approximation
becomes exact as $K$ goes to infinity.  The size $K$ of the sampling
you need \emph{in practice} will depend on the expectations you want
to compute, and the accuracies you need.

As we noted above, in the context of MCMC, we are often sampling or
using some \emph{badly normalized} function $f(x)$.
This function is just the pdf $p(x)$ multiplied by some unknown and
hard-to-compute scalar.
In this case, for our purposes, the conditions on $f(x)$ are that it
be non-negative everywhere and have \emph{finite} integral $Z$
\begin{eqnarray}
0 &\leq& f(x) \quad \mbox{for all $x$}
\\
Z &=& \int f(x)\,\dd x
\quad .
\end{eqnarray}

When the sampling $\setof{x_k}_{k=1}^K$ is of one of these badly
normalized functions $f(z)$---as it usually will be---the
sampling-approximation expectation values are the expectation values
under the properly-normalized corresponding pdf, even though you might
never \emph{learn that normalization}.
When we run MCMC sampling on $f(x)$, a function which differs from a
pdf $p(x)$ by some unknown normalization constant, then the sampling
permits computation of the following kinds of quantities:
\begin{eqnarray}
E[q] &\equiv& \frac{\int q(x)\,f(x)\,\dd x}{\int f(x)\,\dd x}
\\
E[q] &\approx& \frac{1}{K}\,\sum_{k=1}^K q(x_k)
\quad .
\end{eqnarray}
That is, the sampling can be constructed (as we will show below) from
$f(x)$ directly, and it permits you to compute expectation values
without ever requiring you to integrate either the numerator or the
denominator integrals, both of which are generally intractable.\note{As
  we will
  see, the intractability comes from various directions, but one is
  that the dimension of $x$ gets large in most realistic situations.
  Another is that the support of $p(x)$ tends to be, in normal
  inference situations, \emph{far smaller} than the full domain of
  $x$.
  That is, the pdf is at or very close to zero over all but some tiny
  and very hard-to-find part of the space.}

@@

...Is there any way to reconstruct $p(x)$ from the samples
$\setof{x_k}$?  In short, no, but in many cases of interest, sort-of!..

...You can't sample a likelihood!  You can only sample a posterior
pdf!..
why not?..
Give a sense of the difficulties you get into if you fail here; like
discussion with Sven Kreiss on 2013-02-14...

@@

@@

...Autocorrelation time is the premier tool...
How is Autocorrelation time defined?..
How is Autocorrelation time measured?..
Why is it is not a panacea?..

@@

...How do you make it run fast?..

...Acceptance ratio...

...ESJD and etc...

@@

...Gibbs sampling and interior marginalizations...
